The impact of gas bulk rotation on the Lyman-$\alpha$ line

### 2013-08-16 paper/Analisis.tex
some edits

\title[Rotation in the Lyman-$\alpha$ line]{Rotation effects on the
  Lyman-$\alpha$ line morphology in distant galaxies}
\author[Garavito-Camargo, Forero-Romero \& Dijkstra]{
\parbox[t]{\textwidth}{\raggedright 
  Nicolas Garavito-Camargo$^{1}$,
  Jaime E. Forero-Romero$^{1}$ and 
  Mark Dijkstra$^2$
}
\vspace*{6pt}\\
$^{1}$Departamento de F\'{i}sica, Universidad de los Andes, Cra. 1
No. 18A-10, Edificio Ip, Bogot\'a, Colombia \\
$^2$
}
\maketitle

\begin{abstract}
Rotation is present in the gas kinematics of galaxies up to the
highest redshifts. In this paper we present for the first time
radiative transfer calculations that show the impact of rotation on
the morphology of the Lyman $\alpha$ line. To this end we construct
simplified models where a galaxy is modeled as an homogeneous sphere
composed as an homogeneous mixture of dust and hydrogen at a constant
temperature. These spheres have a solid-body rotation with linear
velocities at the surface in the range $0-300$ \kms. We consider
radiation sources both in the center of the rotating cloud and also
homogeneously distributed around the sphere. We find that higher
rotational velocities increase the width of each peak in the outgoing
line profile while it also increases the amount of Lyman alpha photons
escaping in the line center. This trends makes that for high
rotational velocities and large Hydrogen optical depths the double
peak of the line tends to be erased an be replaced by a single peak the
lines center. This is more pronounced for radiation sources
homogeneously distributed. Concerning the escape fraction we find that
rotation does not have any effect, provided that all the sources are
centrally emitted. However in the case of homogeneously emitted
sources we measure an increase of about a factor of $2$ in the escape
fraction for higher rotational velocity values. 
Our work shows clearly that gas rotation has a non negligible impact
on the shape of the Lyman $\alpha$ line. 
\end{abstract}
\begin{keywords}
galaxies: high-redshift - galaxies: star formation - line: formation
\end{keywords}

@@

Due to the resonant nature of the Lyman alpha line, gas kinematics
play an important role shaping its morphology...

There is an extensive literature studying the influence of
outflow/inflow configurations in the shape of the outgoing Lyman-alpha
line...

@@

Describing the kinematics of gas rotation in all generality is a
complex task. There is great variation in the shape of the rotation
curve as observed in HI emission as a function of the distance to the
galaxy center. However there are two features that are observed very
often. First, in the central region the velocity increases
proportional to the radius following the behaviour in a body with
solid rotation. Second, beyond a certain radius the rotation curve
tends to flatten. Furthermore, a thorough observational account of gas
rotation in the redshifts of interest for the study of LAEs ($z>1.0$)
is still missing.

An ab-initio description of realistic rotation curves in simulations
depends on having access to the dynamic evolution the mass components
in the galaxy: stars, gas and dark matter. Such level of realism is
extremely complex to achieve, specially if one wants to get a
systematic description based on statistics of simulated objects.

@@

In the Monte Carlo calculations we follow the propagation of $N_{\gamma}=10^5$
numerical photons through different spherical galaxies, each one
varying in at least one of the following parameters: the maximum
rotational velocity $V_{\rm max}$, hydrogen optical depth $\tau_{H}$,
dust optical depth $\tau_{a}$ and the initial distribution of photons
with respect to the gas. There are in total 60 models with the input
parameters summarized in Table \ref{table:models}.

\begin{table}
\begin{center}
\begin{tabular}{llc}\hline\hline
Physical Parameter (units) & Symbol & Values\\\hline
Velocity (\kms) & $V_{\rm max}$&$0,\ 50,\ 100,\ 200,\ 300$\\
Hydrogen Optical Depth & $\tau_{H} $ & $10^{5},\ 10^{6},\ 10^{7}$\\
Dust Optical Depth & $\tau_{a}$ & $0$,$1$\\
Photons Distributions & & Central, Homogeneous\\\hline\hline
\end{tabular}
\caption{
List of the physical parameters that define the spherical models we
have simulated using Monte Carlo calculations. For each parameter we
vary the values in the range listed in the last column. Takig into
account all the possible combinations we end up with $60$ different
models. } 
\label{table:models}
\end{center}
\end{table}

>>>>

\title[Rotation in the Lyman-$\alpha$ line]{The impact of gas bulk
  rotation on the Lyman-$\alpha$ line} 
\author[Garavito-Camargo, Forero-Romero \& Dijkstra]{
\parbox[t]{\textwidth}{\raggedright 
  Nicolas Garavito-Camargo$^{1}$,
  Jaime E. Forero-Romero$^{1}$ and 
  Mark Dijkstra$^2$
}
\vspace*{6pt}\\
$^{1}$Departamento de F\'{i}sica, Universidad de los Andes, Cra. 1
No. 18A-10, Edificio Ip, Bogot\'a, Colombia \\
$^2$
}
\maketitle

\begin{abstract}
Rotation is present in the gas kinematics of galaxies up to the
highest redshifts. In this paper we present for the first time
radiative transfer calculations that show the impact of gas bulk rotation on
the morphology of the Lyman $\alpha$ line. To this end we model a
galaxy as an homogeneous sphere composed as an homogeneous mixture of
dust and hydrogen at a constant temperature. These spheres have a
solid-body rotation with linear velocities at the surface in the range
$0-300$ \kms and neutral hydrogen optical depths in the range
$\tau_{\rm H}=10^{5}-10^{7}$. We consider radiation sources both in
the center of the rotating cloud and also homogeneously distributed in
the volume. We find that higher rotational velocities increase the
width of each peak in the outgoing line profile while it also
increases the amount of Lyman alpha photons escaping in the line
center. This trends makes that for high rotational velocities and
large Hydrogen optical depths the double peak of the line tends to be
erased an be replaced by a single peak a the line center. This trend
is more pronounced for radiation sources homogeneously
distributed. Concerning the escape fraction we find that rotation does
not have any effect, provided that all the sources are centrally
emitted. However, in the case of homogeneously emittedsources we
measure an increase of about a factor of $2$ in the escape  fraction
for higher rotational velocity values. This work shows that gas bulk
rotation has a non negligible impact on the shape of the Lyman
$\alpha$ line.   
\end{abstract}
\begin{keywords}
galaxies: high-redshift - galaxies: star formation - line: formation
\end{keywords}

@@

[Due to the resonant nature of the Lyman alpha line, gas kinematics
play an important role shaping its morphology.]

[Line morphology is being measured observationally.]

[There is an extensive literature studying the influence of
outflow/inflow configurations in the shape of the outgoing Lyman-alpha
line.]

[Rotation is present in high redshift galaxies.]

@@

Describing the kinematics of gas rotation in all generality is a
complex task, specially at high redshifts where there is still missing
a thorough observational account of rotation in galaxies beyond
$z>1.0$. Furthermore, at low redshifts it has been observed a great
variation in the shape of the rotation curve as observed in HI
emission as a function of the distance to the galaxy center. However
there are two features that are observed very often. First, in the
central region the velocity increases proportional to the radius,
following the behaviour in a body with solid rotation. Second, beyond
a certain radius the rotation curve tends to flatten.

An ab-initio description of realistic rotation curves in simulations
depends on having access to the dynamic evolution of all mass components
in the galaxy: stars, gas and dark matter. Such level of realism is
extremely complex to achieve, specially if one wants to get a
systematic description based on statistics of simulated objects.

@@

In the Monte Carlo calculations we follow the propagation of $N_{\gamma}=10^5$
numerical photons through different spherical galaxies. Each galaxy
model varies at least one of the following parameters: the maximum
rotational velocity $V_{\rm max}$, the hydrogen optical depth $\tau_{H}$,
the dust optical depth $\tau_{a}$ and the initial distribution of photons
with respect to the gas. There are in total 60 models combining all
the input parameters summarized in Table \ref{table:models}.

We have used two independently developed Monte Carlo codes
\citep{CLARA,DijkstraKramer} to perform the calculations of the
non-dusty models. The results we report here are robust in the sense
that they are obtained by both codes.

\begin{table}
\begin{center}
\begin{tabular}{llc}\hline\hline
Physical Parameter (units) & Symbol & Values\\\hline
Velocity (\kms) & $V_{\rm max}$&$0,\ 50,\ 100,\ 200,\ 300$\\
Hydrogen Optical Depth & $\tau_{H} $ & $10^{5},\ 10^{6},\ 10^{7}$\\
Dust Optical Depth & $\tau_{a}$ & $0$,$1$\\
Photons Distributions & & Central, Homogeneous\\\hline\hline
\end{tabular}
\caption{
  List of the physical parameters that define the spherical models 
  imulated using Monte Carlo calculations. For each parameter we
  vary the values in the range listed in the last column. Takig into
  account all the possible combinations we end up with $60$ different
  models. } 
\label{table:models}
\end{center}
\end{table}
